### 2012-05-07 3.1.1.tex
more writing

One of the most salient properties of Turkish is its harmony system, which \citeauthor{Lees1966b} (\citeyear[][35]{Lees1966b}, \citeyear[][284]{Lees1966a}) derives by means of feature spreading rules. Backness harmony is the the most general of these processes.

@@

The rule of \textsc{Backness Harmony} spreads a vowel's \textsc{Back} specification rightward over any intervening consonants onto the next vowel.

@@

\xymatrix@R=24pt@C=24pt{
\txt{V}                                        & \txt{C$_0$} & \txt{V} \\
\txt{[α \textsc{Back}]}\ar@{-}[u]\ar@{--}[urr] \\
}
\end{example}

This rule accounts for a great deal of the suffix allomorphy found in Turkish. For instance, the nominative plural (nom.pl.) suffix is \emph{-ler} when the final root vowel is [$-$\textsc{Back}], and \emph{-lar} when it is [$+$\textsc{Back}], as predicted by Backness Harmony.

@@

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l}
   & \emph{nom.sg.} & \emph{nom.pl.} \\
a. & ip             & ipler          & `rope' & \citep[][216]{Clements1982} \\
   & el             & eller          & `hand'    \\
   & köy            & köyler         & `village' \\
   & yüz            & yüzler         & `face'    \\
   & kız            & kizlar         & `girl'    \\
   & sap            & saplar         & `stalk'   \\
   & son            & sonlar         & `end'     \\
   & pul            & pullar         & `stamp'   \\
b. & neden          & nedenler       & `reason'  & (TELL) \\
   & boğaz          & boğazlar       & `throat'  \\
   & kiler          & kilerler       & `pantry'  \\
   & pelür          & pelürler       & `tissue paper' \\
   & sapık          & sapıklar       & `pervert' \\
\end{tabular}
\end{example}

There are a few complications, however. First, not all polysyllabic roots conform to \textsc{Backness Harmony}. In this case, as can be seen from (\ref{turkexcept}a), suffix vowels generally harmonize with the final root vowel. There is also a small class of nouns, shown in (\ref{turkexcept}b), which exhibit [$-$\textsc{Back}] suffix vowels despite the fact that their final root vowel is [$+$\textsc{Back}].

@@

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l}
   & \emph{nom.sg.} & \emph{nom.pl.} \\
a. & mezar          & mezarlar       & `grave' & \citep{TELL} \\
   & model          & modeller       & `model' \\
   & silah          & silahlar       & `weapon'     \\
   & memur          & memurlar       & `bureaucrat' \\
   & sabun          & sabunlar       & `soap'       \\
b. & saat           & saatler        & `hour, clock' \\
   & harf           & harfler        & `(alphabetic) letter' \\ %& \citep{Goksel2005}
   & etol           & etoller        & `fur stole' \\
\end{tabular}
\end{example}

It is uncontroversial that the disharmonic suffixes of (\ref{turkexcept}b) are no more than very sporadic exceptions to \textsc{Backness Harmony}, root disharmony has ben the subject of much debate. As disharmonic roots still trigger suffix harmony, \citet[][212, 289]{Anderson1974} proposes to distinguish suffix harmony (an alternation) from a sequence structure constraint governing root harmony. This point is echoed by \citet{Iverson1978}.

However, as \citet[][197f.]{Zonneveld1978} notes, the theory of exceptionality proposed in \emph{SPE} (p.~374f.) provides a direct account of suffix harmony in disharmonic roots. \citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule $R$ must be marked [$+$rule $R$] by convention. A root or affix which fails to undergo $R$ despite otherwise matching the structural description is simply said to be marked [$-$rule $R$]. Under this account, exceptionality boils down to failing to match an extended structural description. If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the environment, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.

@@

An alternative account, first proposed by \citet{Clements1982}, uses underspecificatio to derive the non-exceptionality of harmonic roots. The initial-syllable vowel of a Turkish root is contrastively specified for backness (e.g., \emph{kül} `ash' vs.  \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'); harmonic roots are ones in which non-initial syllable vowels cannot contrast for this feature, and \citeauthor{Clements1982} proposes to leave them underspecified. Only in disharmonic roots does \textsc{Back} need to be specified for non-initial syllable vowels. 
%This is exemplified below in (\ref{spec}).
%(e.g., \emph{deve} `camel' vs. \emph{deva} `medicine', \emph{sene} `year' vs. \emph{sena} `praise').

%There is some further evidence that individual vowels may differ in specification for this feature even within individual roots or affixes. For instance, the present continuous suffix has harmony-determined allomorphs \emph{-iyor}, \emph{-üyor}, \emph{-ıyor}, \emph{-uyor}, but the \emph{o} of the suffix is invariant. A similar situation might obtain in Turkish roots.

\begin{example}
\label{spec}
\textsc{Back} specification for (dis)harmonic roots:

\vspace{0.5\baselineskip}
\xymatrix@R=24pt@C=24pt{
\txt{a.} & \txt{harmonic root:~~~~} & \txt{C} & \txt{V} & \txt{C} & \txt{V} \\
&   &    & \txt{[$-$\textsc{Back}]}\ar@{-}[u]\ar@{--}[urr] \\
\txt{b.} & \txt{disharmonic root:} & \txt{C} & \txt{V} & \txt{C} & \txt{V} \\
    &    &         & \txt{[$-$\textsc{Back}]}\ar@{-}[u] & & \txt{[$+$\textsc{Back}]}\ar@{-}[u]
}
\end{example}

\noindent
Harmonizing suffix vowels will also be underspecified for \textsc{Back}. Of course, \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, one option being the use of a a \textsc{Structure Preservation} condition \citep{Kiparsky1985}. Unfortunately, this detail reintroduces the duplication alluded to above. Any condition on the rule of \textsc{Backness Harmony} which prevents overwriting entails that it has no control over the distribution of harmonic and disharmonic roots, and thus fails to account for predominance of harmonic roots.

\citet{Clements1982} and \citet{Kaun1999}

\footnote{Thanks to Kie Zuraw for bringing this data to my attention.}

the epenthetic vowel is always [$+$\textsc{High}], and when the first consonant of the cluster is non-dorsal, it shows backness harmony

@@

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l l}
a. & spiker  & \alt{} & sipiker  & `announcer' \\
   & fren    & \alt{} & firen    & `break'     \\
   & trablus & \alt{} & tırablus & `Tripoli'   \\
   & brom    & \alt{} & burom    & `bromide'   \\
   & prusya  & \alt{} & purusya  & `Prussia'   \\
b. & kral    & \alt{} & kıral    & `king'      \\
   & grup    & \alt{} & gurup    & `group'     \\
c. & grip    & \alt{} & gırip    & `grippe'    \\ % unexpectedly back
   & kredi   & \alt{} & kıredi   & `credit'    \\
\end{tabular}
\end{example}

\subsubsection{Psycholinguistic evidence}

The evidence from alternations leaves open the question of whether speakers internalize a generalization regarding the tendency of roots to conform to \textsc{Backness Harmony}. One piece of evidence that speaks in favor of root-internal harmony comes from a language game discussed by \citet{Harrison2001}.\footnote{Thanks to Bert Vaux for bringing this study to my attention.} This game is not indigenous to Turkish, but it corresponds to a morphological process native to the related language Tuvan, in which it conveys a sense of ``vagueness or jocularity'', and \citeauthor{Harrison2001} report that it can be quickly taught to even young Turkish speakers. The game consists of reduplication of the base and replacement of the first vowel in the reduplicant with \emph{a} or \emph{u}.

In both Tuvan and Turkish, the unchanged reduplicant vowel is also affected. Reduplication interacts with root harmony in both Tuvan and Turkish. When the base is harmonically [$-$\textsc{Back}], the insertion of a [$+$\textsc{Back}] results in what \citeauthor{Harrison2001} call ``reharmonization'' (\ref{redupgame}a).

@@

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l}
a. & kibrit & kibrit-kabrıt & `match'    \\
   & bütün  & bütün-batın   & `whole'    \\
b. & mali   & mali-muli     & `Mali'     \\
   & butik  & butik-batik   & `boutique' \\
\end{tabular}
\end{example}

\noindent
Under the underspecification hypothesis, non-initial harmonic vowels lack an underlying \textsc{Back} feature, so it comes as no surprise that changing the \textsc{Back} specification of the root-initial vowel results in reharmonization. And the full specification of disharmonic roots correctly predicts that they will be exempt from reharmonization (\ref{redupgame}b), which is also borne out in Tuvan and in an unrelated Finnish language game \citep{Campbell1986}.

A number of studies have found that speakers of Finnish use disharmony in speech processing experiments.\footnote{Thanks to Charles Yang for pointing out the relevance of these studies to me.} \citet{Suomi1997} and \citet{Vroomen1998} generate nonce trisyllabic words by adding a monosyllabic pseudo-prefix to real and nonce disyllabic words, all of which are harmonic for the feature \textsc{Back}. These stimuli are auditorily presented to subjects who are asked to press a button when the nonce trisyllable ends with a target nonce disyllable, or a real disyllabic word. Speakers are quicker to press the button when the prefix and disyllabic word disagree for \textsc{Back}. These results suggest that speakers are attuned to the fact that disharmonic transisitions are good predictors of word boundaries. If speakers have also internalized the converse generalization, that harmonic transistions are more likely to be root-internal, then there is additional evidence that harmony is active not just in Finnish affix alternations but also in roots.

\citet{Kabak2010} report that Turkish \textsc{Backness Harmony} has the same effect on word-spotting as it does in Finnish: speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by the pseudo-prefix \emph{gölü-}, a disharmonic transistion, than when it is preceded by the pseudo-prefix \emph{golu-}, a harmonic transition. \citeauthor{Kabak2010} find that effect of harmony does not obtain for speakers of French, a language which lacks vowel harmony. As in Finnish, the results imply speakers have internalized the predominance of root-internal harmony.

>>>>

\citeauthor{Lees1966b} (\citeyear[][35]{Lees1966b}, \citeyear[][284]{Lees1966a}) models the Turkish vowel harmony systems by means of three feature spreading rules. Backness harmony is the the most general of these processes.

@@

\textsc{Backness Harmony} spreads the backness specification of a vowel rightward ignoring any intervening consonants.

@@

\xymatrix@R=24pt@C=24pt{
\txt{V}                                        & \txt{C$_0$~~~~~} & \txt{V} & \txt{(condition: rightward application)} \\
\txt{[α \textsc{Back}]}\ar@{-}[u]\ar@{--}[urr] \\
}
\end{example}

This rule accounts for the tendency of polysyllablic roots to either contain all [$+$\textsc{Back}] or all [$-$\textsc{Back}];
%, as can be seen in the nominative singulars in (\ref{turknom}b); 
this tendency is quantified in \S\ref{lexstats} below. \textsc{Backness Harmony} also accounts for the allomorphs of several inflectional suffixes. For instance, the nominative plural (nom.pl.) suffix is \emph{-ler}, when the final root vowel is [$-$\textsc{Back}], and \emph{-lar} when it is [$+$\textsc{Back}].

@@

%\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l}
   & \emph{nom.sg.} & \emph{nom.pl.} \\
a. & ip             & ipler          & `rope' & \citep[][216]{Clements1982} \\
   & el             & eller          & `hand'    \\
   & köy            & köyler         & `village' \\
   & yüz            & yüzler         & `face'    \\
   & kız            & kizlar         & `girl'    \\
   & sap            & saplar         & `stalk'   \\
   & son            & sonlar         & `end'     \\
   & pul            & pullar         & `stamp'   \\
b. & neden          & nedenler       & `reason'  & (TELL) \\ % front
   & kiler          & kilerler       & `pantry'           \\ % front
   & pelür          & pelürler       & `tissue paper'     \\ % back
   & boğaz          & boğazlar       & `throat'           \\ % back
   & sapık          & sapıklar       & `pervert'          \\ % back
\end{tabular}
\end{example}

There are a few complications, however. First, not all polysyllabic roots conform to \textsc{Backness Harmony}. In this case, as shown in (\ref{turkexcept}a), suffix vowels generally harmonize with the final root vowel. There is also a very small class of nouns, shown in (\ref{turkexcept}b), which take \emph{-ler} despite the fact that their final root vowel is [$+$\textsc{Back}].

@@

%\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l}
   & \emph{nom.sg.} & \emph{nom.pl.} \\
a. & mezar          & mezarlar       & `grave' & \citep{TELL} \\
   & model          & modeller       & `model' \\
   & silah          & silahlar       & `weapon'     \\
   & memur          & memurlar       & `bureaucrat' \\
   & sabun          & sabunlar       & `soap'       \\
b. & saat           & saatler        & `hour, clock' \\
   & harf           & harfler        & `(alphabetic) letter' \\ %& \citep{Goksel2005}
   & etol           & etoller        & `fur stole' \\
\end{tabular}
\end{example}

While it is uncontroversial that the disharmonic suffixes of (\ref{turkexcept}b) are no more than very sporadic exceptions to \textsc{Backness Harmony}, root disharmony has ben the subject of much debate. As disharmonic roots still trigger suffix harmony, \citet[][212, 289]{Anderson1974} and \citet{Iverson1978} propose to separate suffix harmony, an alternation, from a sequence structure constraint governing root harmony.

The disadvantage of this account is that it introduces a sort of ``duplication'' (in the sense of \citealt{Kenstowicz1977}) of sequence structure and phonological generalizations, differning only in their patterns of exceptionality.
%\citep{Kisseberth1970b} (see also \citealp[][401]{Stanley1967} and \emph{SPE}:382)
However, \citet[][197f.]{Zonneveld1978} observes that the theory of exceptionality proposed in \emph{SPE}
% (p.~374f.)
can account for suffix harmony in disharmonic roots. \citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule \emph{R} must be marked [$+$\emph{R}] by convention. A root or affix which fails to undergo \emph{R} despite otherwise matching the structural description is simply said to be marked [$-$\emph{R}]. In other words, a form is never truly an ``exception'': it simply fails to match an extended structural description including the rule feature. If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the trigger, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.\footnote{The definition of ``target of a rule'' is not obvious in autosegmental rules, but any useful formalization of this notion should, for example, identify a harmonic suffix vowel as the target of \textsc{Backness Harmony}.} With this in place, \textsc{Backness Harmony} is sufficient to account for both root and suffix harmony.

@@

Before moving on, it is necessary to dispense with an alternative analysis originally proposed by \citet{Clements1982}. Root vowels exhibit a robust contrast for backness (e.g., \emph{kül} `ash' vs.  \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'), whereas harmonic roots are those in which the backness of any remaining vowels is predictable. \citeauthor{Clements1982} propose that these vowels are underspecified for backness, whereas the non-initial vowels of disharmonic roots are fully specified. This is schematized below.

%This is exemplified below in (\ref{spec}).
%(e.g., \emph{deve} `camel' vs. \emph{deva} `medicine', \emph{sene} `year' vs. \emph{sena} `praise'). 
%There is some further evidence that individual vowels may differ in specification for this feature even within individual roots or affixes. For instance, the present continuous suffix has harmony-determined allomorphs \emph{-iyor}, \emph{-üyor}, \emph{-ıyor}, \emph{-uyor}, but the \emph{o} of the suffix is invariant. A similar situation might obtain in Turkish roots.

\begin{example}
\label{spec}
Underspecification in harmonic roots (after \citealp{Clements1982}):

%\vspace{0.5\baselineskip}
\xymatrix@R=24pt@C=24pt{
\txt{a.} & \txt{harmonic root:~~~~} & \txt{C} & \txt{V} & \txt{C} & \txt{V} \\
&   &    & \txt{[$-$\textsc{Back}]}\ar@{-}[u]\ar@{--}[urr] \\
\txt{b.} & \txt{disharmonic root:} & \txt{C} & \txt{V} & \txt{C} & \txt{V} \\
    &    &         & \txt{[$-$\textsc{Back}]}\ar@{-}[u] & & \txt{[$+$\textsc{Back}]}\ar@{-}[u]
}
\end{example}

Harmonizing suffix vowels will also be underspecified for backness.

Two additional details are needed to complete this analysis. First, there must be some kind of constraint which prevents speakers from positing redundant backness specifications for non-initial vowels in harmonic roots, a constraint on identical adjacent specifications in underlying representations, the lexical ``obligatory contour principle'' \citep[][OCP]{Leben1973}, on surface tonal representations \citep{Goldsmith1976}, or both (\citealp{Leben1978}, \citealp{McCarthy1986}; see \citealp{Odden1986,Odden1988} for criticism). Second, \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, one option being the use of a structure preservation condition \citep{Kiparsky1985}. However, any condition which prevents \textsc{Backness Harmony} from overwriting underlying backness specifications will reintroduce the duplication of sequence structure and phonological generalizations; under this analysis, disharmonic roots are no longer exceptional, despite considerable evidence (reviewed below) that they are formally marked in Turkish.\footnote{On the other hand, if one interprets ``harmony'', that is, a single backness specification per root, as the learner's default (as \citealp{Odden1986} proposes for the tonal obligatory contour principle), this account is a mere notational variant of the rule exceptionality account.}
%\footnote{Kie Zuraw (p.c.) suggests that the markedness of disharmonic roots might be attributed to the presence of more structure, namely, multiple specifications for backness.}
I reject the underspecification analysis on these grounds. Rather, I assume a unitary phonological rule of \textsc{Backness Harmony} and assign [$-$\textsc{Backness Harmony}] to disharmonic roots.

\subsubsection{External evidence}
\label{backharmexternal}

While suffix harmony can be inferred from alternations, no evidence has yet been presented to show that speakers internalize in any way tendency for roots to exhibit backness harmony.

%\begin{quote}
%\ldots{}these generalizations\ldots{}have no observable consequences in the course of the normal use of the language. \citep[][320]{Zimmer1969}
%\end{quote}

Indeed, \citeauthor{PE} adopt the null hypothesis that speakers do not internalize any generalizations about possible or likely sequences in underlying representations.

\begin{quote}
Even if we, as linguists, find some generalizations in our description of the lexicon, there is no reason to posit these generalizations as part of the speaker's knowledge of their language, since they are computationally inert and thus irrelevant to the input-output mapping that the grammar is responsible for. \citep[][18]{PE}
\end{quote}

Several independent pieces of external evidence argue that this is not the case for root harmony in Turkish. 
The adaptation of non-native word-initial onset clusters, discussed by \citet{Clements1982} and \citet{Kaun1999}, is one such case.\footnote{Thanks to Kie Zuraw for bringing this data to my attention.} While some speakers are said to be capable of pronouncing these non-native clusters, in normal speech the cluster is resolved by epenthesis of a [$+$\textsc{High}] vowel. In most cases, the epenthetic vowel matches the following root vowel for backness, suggesting that adapt foreign words in a way that conforms to the root harmony generalization.

%the epenthetic vowel is always [$+$\textsc{High}],
%and in general, it conforms to \textsc{Backness Harmony}.
%For reasons that are 
%and when the first consonant of the cluster is non-dorsal, it shows backness harmony

@@

%\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l l}
a. & spiker  & \alt{} & sipiker  & `announcer' \\
   & fren    & \alt{} & firen    & `break'     \\
b. & brom    & \alt{} & burom    & `bromide'   \\
   & prusya  & \alt{} & purusya  & `Prussia'   \\
%b. & grip    & \alt{} & gırip    & `grippe'    \\ % unexpectedly back
%   & kredi   & \alt{} & kıredi   & `credit'    \\
%   & trablus & \alt{} & tırablus & `Tripoli'   \\
%b. & kral    & \alt{} & kıral    & `king'      \\
%   & grup    & \alt{} & gurup    & `group'     \\
\end{tabular}
\end{example}

Similar evidence comes from a language game discussed by \citet{Harrison2001}.\footnote{Thanks to Bert Vaux for bringing this study to my attention.} This game is not indigenous to Turkish, but is identical to a morphological process in the related language Tuvan, in which it conveys a sense of ``vagueness or jocularity'', and \citeauthor{Harrison2001} report that it can be quickly taught to Turkish speakers. The game consists of reduplication of the base and replacement of the first vowel in the reduplicant with the [$+$\textsc{Back}] vowels \emph{a} or \emph{u}. Both in the native Tuvan process and in the Turkish game, the second vowel of the reduplicant (shown in braces below) is affected by this process: when the base is harmonically [$-$\textsc{Back}], the insertion of a [$+$\textsc{Back}] results in what \citeauthor{Harrison2001} call ``reharmonization'', shown in  (\ref{redupgame}a).

@@

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l}
a. & kibrit & kibrit-\{kabrıt\} & `match'    \\
   & bütün  & bütün-\{batın\}   & `whole'    \\
b. & mali   & mali-\{muli\}     & `Mali'     \\
   & butik  & butik-\{batik\}   & `boutique' \\
\end{tabular}
\end{example}

\noindent \citeauthor{Harrison2001} analyze this in terms of the \citeauthor{Clements1982} underspecification analysis, but it is equally consistent with full specification. Reharmonization is simply the spread of the backness specification of the \emph{a} or \emph{u} in the reduplicant. On the other hand, (\ref{redupgame}b) shows that this does not obtain for disharmonic roots; presumably, the [$-$\textsc{Backness Harmony}] exception feature is copied under reduplication as well. Similar results obtain both in Tuvan and in an unrelated Finnish language game \citep{Campbell1986}.

A number of studies on Finnish suggest that speakers of that language, which has a vowel harmony system similar to that of Turkish, make use of (dis)harmony to processing running speech.\footnote{Thanks to Charles Yang for bringing these studies to my attention.} \citet{Suomi1997} and \citet{Vroomen1998} generate nonce trisyllabic words by adding a monosyllabic pseudo-prefix to real and nonce disyllabic words, all of which conform to backness harmony. These stimuli are auditorily presented to subjects who are asked to press a button when the nonce trisyllable ends with a target nonce disyllable, or a real disyllabic word. Speakers are quicker to press the button when the prefix and disyllabic word have different backness specifications. These results suggest that speakers are attuned to the fact that disharmonic transisitions are good predictors of word boundaries. If speakers have also internalized the converse generalization, that harmonic transistions are more likely to be root-internal, then there is additional evidence that harmony is active not just in Finnish affix alternations but also in roots.

\citet{Kabak2010} report that Turkish \textsc{Backness Harmony} has the same effect in word-spotting tasks as it does in Finnish: speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by the pseudo-prefix \emph{gölü-}, a disharmonic transistion, than when it is preceded by the pseudo-prefix \emph{golu-}, a harmonic transition. \citeauthor{Kabak2010} find that effect of harmony does not obtain for speakers of French, a language which lacks vowel harmony. As in Finnish, the results imply speakers have internalized the predominance of root-internal harmony.

### 2012-05-07 3.1.2.tex
more writing

The rule of roundness harmony and the data that motivates it overlaps with the preceding evidence for \textsc{Backness Harmony}.

@@

The environment for roundness harmony differs from \textsc{Backness Harmony} in that it requiers the target to be [$+$\textsc{High}].

@@

\xymatrix@R=24pt@C=24pt{
\txt{[α \textsc{Round}]}\ar@{-}[d]\ar@{..}[drr] &             & \txt{[$+$\textsc{High}]} \\
\txt{V}                                         & \txt{C$_0$} & \txt{V}\ar@{-}[u] \\
}
\end{example}

@@

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l l l}
   & \emph{nom.sg.} & \emph{dat.sg.} & \emph{gen.sg.}  \\
a. & ip             & ipi            & ipin           & `rope' & (CS:216) \\
   & el             & eli            & elin           & `hand'    \\
   & kız            & kızı           & kızın          & `girl'    \\
   & sap            & sapı           & sapın          & `stalk'   \\
   & yüz            & yüzü           & yüzün          & `face'    \\
   & köy            & köyü           & köyün          & `village' \\
   & pul            & pulu           & pulun          & `stamp'   \\
   & son            & sonu           & sonun          & `end'     \\
b. & boğaz          & boğazı         & boğazın        & `throat'  & (TELL) \\
   & pelür          & pelürü         & pelürün        & `tissue paper' \\
   & döviz          & dövizi         & dövizin        & `currency' \\
   & yamuk          & yamuğu         & yamuğun        & `trapezoid' \\
   & ümit           & ümiti          & ümitin         & `hope'     \\
\end{tabular}
\end{example}

\noindent
Much as was the case for \textsc{Backness Harmony}, there are disharmonic roots which participate in suffix harmony. Once again, underspecification of harmonic roots and full specification of disharmonic roots allows for the use of the \emph{SPE} exception convention.

\subsubsection{Psycholinguistic evidence}

The only psycholinguistic evidence for \textsc{Roundness Harmony} comes from \citeauthor{Harrison2001}'s language game (\ref{redupgame}). As shown above, the second vowel in the reduplicated form of \emph{bütün} `whole' is \emph{bütün-batın}. The second vowel in the base, \emph{ü}, is reharmonized as [$+$\textsc{Back}, $-$\textsc{Round}], indicating that \textsc{Roundness Harmony} also participates in reharmonization.

>>>>

Roundness harmony is a closely related process with a somewhat narrower scope.

@@

Roundness harmony is quite similar to \textsc{Backness Harmony}, with the additional stipulation that the target must be [$+$\textsc{High}].

@@

\xymatrix@R=24pt@C=24pt{
\txt{[α \textsc{Round}]}\ar@{-}[d]\ar@{--}[drr] &             & \txt{[$+$\textsc{High}]} \\
\txt{V}                                         & \txt{C$_0$} & \txt{V}\ar@{-}[u] \\
}
\end{example}

@@

%\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l l l}
   & \emph{nom.sg.} & \emph{dat.sg.} & \emph{gen.sg.}  \\
a. & ip             & ipi            & ipin           & `rope' & \citep[][216]{Clements1982} \\
   & el             & eli            & elin           & `hand'    \\
   & kız            & kızı           & kızın          & `girl'    \\
   & sap            & sapı           & sapın          & `stalk'   \\
   & yüz            & yüzü           & yüzün          & `face'    \\
   & köy            & köyü           & köyün          & `village' \\
   & pul            & pulu           & pulun          & `stamp'   \\
   & son            & sonu           & sonun          & `end'     \\
b. & boğaz          & boğazı         & boğazın        & `throat'  & (TELL) \\
   & pelür          & pelürü         & pelürün        & `tissue paper' \\
   & döviz          & dövizi         & dövizin        & `currency' \\
   & yamuk          & yamuğu         & yamuğun        & `trapezoid' \\
   & ümit           & ümiti          & ümitin         & `hope'     \\
\end{tabular}
\end{example}

\noindent As was the case for \textsc{Backness Harmony}, disharmonic roots are found, and the final vowel of disharmonic roots triggesr suffix harmony.

\subsubsection{External evidence}

Though few studies have directly sought external evidence for speaker's awareness of root \textsc{Roundness Harmony}, two of the findings reviewed in \S\ref{backharmexternal} above speak in support of this claim. First, the epenthetic vowel in non-native word-initial onsets
like \emph{trablus} \alt{} \emph{tırablus} `Tripoli' conform also to \textsc{Roundness Harmony}.
%\citealp[][247]{Clements1982}). 
Second, \textsc{Roundness Harmony} participates in reharmonization in \citeauthor{Harrison2001}'s language game: the reduplicated \emph{bütün-batın} exhibits replacement of the second \emph{ü}, a [$+$\textsc{Round}] vowel, with [$-$\textsc{Round}] \emph{ı}.

### 2012-05-07 3.1.3.tex
more writing

The final rule

@@

\xymatrix@R=24pt@C=8pt{
\txt{[$-$\textsc{Round}]} &                                         & \txt{[$-$\textsc{High}]} & \txt{[$+$\textsc{Labial}]}    & \txt{[$+$\textsc{High}]}\ar@{-}[dr] &         & \txt{[$+$\textsc{Round}]}\ar@{..}[dl] \\
                         & \txt{V}\ar@{-}[ul]\ar@{-}[ur]\ar@{-}[d] &                           & \txt{C$_0$ C C$_0$}\ar@{-}[u] &                                      & \txt{V} & \\
                         & \txt{[$+$\textsc{Back}]}\ar@{-}[urrrr] 
}
\end{example}

>>>>

%\begin{quote}
%\ldots{}these generalizations\ldots{}have no observable consequences in the course of the normal use of the language. \citep[][320]{Zimmer1969}
%\end{quote}

@@

\xymatrix@R=24pt@C=8pt{
\txt{[$-$\textsc{Round}]} &                                         & \txt{[$-$\textsc{High}]} & \txt{[$+$\textsc{Labial}]}    & \txt{[$+$\textsc{High}]}\ar@{-}[dr] &         & \txt{[$+$\textsc{Round}]}\ar@{--}[dl] \\
                         & \txt{V}\ar@{-}[ul]\ar@{-}[ur]\ar@{-}[d] &                           & \txt{C$_0$ C C$_0$}\ar@{-}[u] &                                      & \txt{V} & \\
                         & \txt{[$+$\textsc{Back}]}\ar@{-}[urrrr] 
}
\end{example}

### 2012-05-07 3.1.tex
more writing

\citet{Lees1966a,Lees1966b} proposes three constraints on Turkish vowel sequences 
which have been the attention of many subsequent studies on vowel harmony and on sequence structure constraints. In this section, these constraints are formalized, and where possible, related to phonological alternations in Turkish and to various sources of external evidence. The following feature specification for the Turkish vowels is assumed throughout.

>>>>

\citet{Lees1966a,Lees1966b} proposes three constraints on Turkish vowel sequences which have been the focus of many subsequent studies. In this section, these constraints are formalized, and where possible, related to phonological alternations in Turkish and to evidence that speakers have internalized these restrictions. The following feature specification for the Turkish vowels is assumed throughout.

### 2012-05-07 3.2.2.tex
another day at the office
more writing

There is some disagreement in the literature about just what \citet{Zimmer1969} finds regarding \textsc{Labial Attraction}.

\citet{Zimmer1969}
(see also \citet[][4]{Zuraw2000})
claims however, Zimmer's evidence suggests that the speakers did internalize a version of the labial attraction generalization.

@@

since the opposing are all broader versions than the original formulation by \citeauthor{Lees1966a}.

\citet{Ito1993} argue that the data shows that it holds of native words only

@@

($\gamma = 0.597$, $p = 5.2$\e{-29})

@@

@@

@@

($\gamma = 0.087$, $p = 0.101$)

>>>>

There is some disagreement in the literature about just what \citeauthor{Zimmer1969}'s study reveals about \textsc{Labial Attraction}.

zimmer
clements
\citet{Ito1993} argue that the data shows that it holds of native words only
\citet{Inkelas2001} refute this point

\citet{Zuraw2000}:4

@@

%since the opposing are all broader versions than the original formulation by \citeauthor{Lees1966a}.

@@

@@

$\gamma = 0.597$, $p = 5.2$\e{-29}

@@

$\gamma = 0.641$, $p = 7.7$\e{-27}

@@

$\gamma = 0.087$, $p = 0.101$
